Highly Available Transactions: Virtues and Limitations

### 2013-03-29 writeup/intro.tex
finishing pdb edit pass

Requirements for high availability, predictable low latency, and
simplicity of coordination protocols have led many recent distributed
database designers to renounce several guarantees provided by
traditional database systems~\cite{bigtable, pnuts, dynamo}. In
particular, many modern systems forgo providing ACID guarantees, which
often provide single-system image programmability but suffer from
unavailability and often expensive coordination
costs~\cite{davidson-survey}. In particular, the July 2000
announcement of the CAP Theorem~\cite{gilbert-cap}, which states that
a highly available system cannot provide recency guarantees in the
presence of network partitions, proved a rallying call for a
generation of scalable but semantically weak systems. Over a decade
later, we observe a dearth of highly available systems providing
general-purpose multi-operation \textit{transactions} over multiple
data items with useful semantics.

While serializability, the gold standard of traditional ACID
transactions, is not achievable with high availability for general
read-write transactions~\cite{davidson-survey}, databases have a long
tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
concurrency and overall performance benefits, weak isolation models
are widely employed, overwhelmingly as the default setting in today's
traditional ``ACID'' and up-and-coming ``NewSQL'' databases and often
as the maximum guarantee offered (Section~\ref{sec:modernacid}). While
this weak isolation does not provide serializability for
general-purpose transactions, it is apparently strong enough to
deliver acceptable behavior to many application programmers and is
substantially stronger than the typical semantics of current highly
available systems. This raises the question: which of these semantics
can we provide while remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
is poorly understood. We have a strong understanding of weak isolation
in the single-server context from which it originated~\cite{adya,
  ansicritique, gray-isolation} as well as several techniques for
providing distributed
serializability~\cite{bernstein-concurrency}. However, the current
literature does not provide guidance as to the operation of weakly
consistent transactional guarantees in a highly available distributed
context. Accordingly, in this work, we consider the problem of
providing Highly Available Transactions (HATs), or semantic guarantees
regarding multiple operations operating on multiple data items that
are achievable with high availability and low latency, albeit with
some cost to achievable semantics. While our preliminary results
indicate that some models are achievable~\cite{hat-hotos}, the
literature lacks a deep understanding of what models are available,
which are not, and when this matters.

This work attempts to bridge the gap between ACID properties and
highly available systems. We introduce a hierarchy of consistency
models and rigorously taxonomize their availability characteristics;
in doing so, we unify previous work on ACID properties~\cite{adya},
session guarantees~\cite{sessionguarantees}, and traditional
disributed ``register'' semantics~\cite{herlihy-art}. By identifying
the availability cost of a wide range of popular consistency
guarantees, this work allows developers to weigh the costs and
benefits of relying on these guarantees in their applications. While
several guarantees, like serializability and Snapshot Isolation, are
unachievable due to update conflict avoidance (preventing Lost Update
and Write Skew phenomena), many other ACID guarantees are HAT
compliant, such as providing Read Committed isolation, preventing
Fuzzy Read phenomena, and ensuring atomicity of transactional
updates. Several other guarantees, like reading one's writes, fall
into a new availability class---``sticky'' high availability---which
provides an intermediate ground between strict high availability and
majority availability.

This taxonomy allows us to weigh the costs and benefits of a wide
range of semantic guarantees, but which guarantees are worthwhile in
practice? We attempt to objectively study the virtues and limitations
of both classic and HAT systems by performing an extensive survey of
practitioner accounts and research literature, performing experimental
analyses on modern cloud infrastructure, and analyzing representative
applications for their semantic requirements. While our results are
not definitive for all applications, they suggest that HATs offer a
one to three order of magnitude latency decrease compared to
traditional protocols and provide acceptable semantics for a wide
range of programs, particularly those with monotonic logic and
commutative updates~\cite{calm, blooml, crdt}. HAT systems can also
provide arbitrary foreign key constraints on multi-item updates and
sometimes provide uniqeness guarantees. However, particularly for
programs with non-monotonic logic, HATs may fall short, requiring more
coordination-intensive mechanisms. Concisely, we attempt to understand
\textit{what benefits do Highly Available Transactions offer and what
  is their cost?}

In this paper, we make the following contributions:

\begin{itemize}
\item We model high availability in a transactional environment,
  including traditional and sticky high availability.

@@

\item We analyze existing concurrency control algorithms, perform a
  case-study of an existing transactional application, and
  experimentally evaluate a HAT database prototype.
\end{itemize}

>>>>

Requirements for high availability, predictable low latency, and
simplicity of coordination protocols have led many recent distributed
database designers to renounce several properties provided by
traditional database systems~\cite{bigtable, pnuts, dynamo}. In
particular, many modern systems forgo providing ACID guarantees, which
often provide single-system image programmability but traditionally
suffer from unavailability and often expensive coordination
costs~\cite{davidson-survey}. In particular, the July 2000
announcement of the CAP Theorem~\cite{gilbert-cap}, which states that
a highly available system cannot provide recency guarantees in the
presence of network partitions, proved a rallying call for a
generation of scalable but semantically weak system designs. Over a
decade later, we observe a dearth of highly available systems
providing general-purpose multi-operation \textit{transactions} over
multiple data items with useful semantics.

While serializability, the gold standard of traditional ACID
transactions, is not achievable with high availability for general
read-write transactions~\cite{davidson-survey}, database systems have
a long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
concurrency and overall performance benefits, weak isolation models
are widely employed, overwhelmingly as the default setting in today's
traditional ``ACID'' and up-and-coming ``NewSQL'' databases and often
as the maximum guarantee offered (Section~\ref{sec:modernacid}). While
this weak isolation does not provide serializability for
general-purpose transactions, it is apparently strong enough to
deliver acceptable behavior to many application programmers and is
substantially stronger than the typical semantics of current highly
available systems. This raises the question: which of these semantics
can be provided while remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
is poorly understood. We have a strong understanding of weak isolation
in the single-server context from which it originated~\cite{adya,
  ansicritique, gray-isolation} as well as several techniques for
providing distributed serializability~\cite{bernstein-concurrency,
  spanner, granola, daudjee-session, calvin}. However, the current
literature does not provide guidance as to the operation of weakly
consistent transactional guarantees in a highly available distributed
context. Accordingly, in this work, we consider the problem of
providing Highly Available Transactions (HATs), or semantic guarantees
regarding multiple operations operating on multiple data items that
are achievable with high availability and low latency, albeit with
some cost to achievable semantics. While our prior, preliminary
results indicated that some models are achievable~\cite{hat-hotos},
the literature lacks a deep understanding of what models are
available, which are not, and when this distinction matters.

This work attempts to bridge the gap between ACID properties and
highly available systems. We introduce a hierarchy of consistency
models and rigorously taxonomize their availability characteristics;
in doing so, we unify previous work on ACID properties~\cite{adya},
session guarantees~\cite{sessionguarantees}, and traditional
disributed ``register'' semantics~\cite{herlihy-art}. By identifying
the availability cost of a wide range of popular consistency
guarantees, this work allows developers to weigh the costs and
benefits of relying on particular guarantees in their applications. While
several models, like serializability and Snapshot Isolation, are
unachievable due to update conflict avoidance (preventing Lost Update
and Write Skew phenomena), many other ACID guarantees are HAT
compliant, such as providing Read Committed isolation, preventing
Fuzzy Read phenomena, and ensuring atomicity of transactional
updates. Several other guarantees, like reading one's writes, fall
into a new availability class---``sticky'' high availability---which
provides an intermediate ground between strict high availability and
majority availability.

This taxonomy allows us to weigh the costs and benefits of a wide
range of semantic guarantees, but which guarantees are worthwhile in
practice? We attempt to objectively study the virtues and limitations
of both classic and HAT systems by surveying practitioner accounts and
research literature, performing experimental analyses on modern cloud
infrastructure, and analyzing representative applications for their
semantic requirements. While our results are not definitive for all
applications, they suggest that HATs offer a one to three order of
magnitude latency decrease compared to traditional protocols and can
provide acceptable semantics for a wide range of programs,
particularly those with monotonic logic and commutative
updates~\cite{calm, blooml, crdt}. HAT systems can also enforce
arbitrary foreign key constraints on multi-item updates and sometimes
provide limited uniqueness guarantees. However, particularly for
programs with non-monotonic logic, HATs may fall short, requiring more
coordination-intensive, unavailable mechanisms. Concisely, we attempt
to understand \textit{what benefits do Highly Available Transactions
  offer and what is their cost?}

In this paper, we make the following contributions:
\begin{myitemize}
\item We model high availability in a transactional environment,
  including traditional and sticky high availability.

@@

\item We analyze existing concurrency control algorithms, perform a
  case-study of an existing transactional application, and
  briefly evaluate a HAT database prototype.
\end{myitemize}

### 2013-03-29 writeup/model.tex
finishing pdb edit pass

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response as the (set of) server(s) does not need to communicate with
any others. If servers are partitioned from one another, they do not
need to stall in order to provide a ``safe'' query response. This lack
of fast-path coordination also means that a highly available system
also provides low latency~\cite{abadi-pacelc}. However, to properly
describe whether a \textit{transactional} system is highly available,
we need both a way to describe what servers a client must contact as
well as what kinds of responses a server can provide, especially given
the possibility of aborts.

@@

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between a
  fully replicated system, in which all servers are replicas for all
  data items, and a partially replicated system, in which at least one
  server acts as a replica for only a subset of all data items. For
  generality, and, for applicability to many modern ``sharded'' or
  ``partitioned'' data storage systems like Dynamo, we consider
  partially replicated systems.} a server that can handle an
operation for a given data item is called a \textit{replica} for that
item. This is a useful definition: if a client can contact at least
one replica for an item on which it wishes to perform an operation,
then the replica will provide a response to the client. However, there
are several other cases we should consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of operations will persist in the face of
server failures, clients may wish to wait until operations reach more
than one server before they consider operations completed. However,
according to our definition of high availability, server-fault
tolerance is unavailable. To allow flexibility in the number of
servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing high availability ($1$-availability), a system with
$2$-availability will require more messaging and possibly higher
latency and may experience more unsuccessful operations in the
presence of network and server failures.  Additionally, as is commmon
in standard consensus algorithms like Paxos, clients may want to
contact a majority of $N$ servers in a system. This \textit{majority
  availability} (a special case of $K$-availability with $K=\lceil
\frac{N}{2} \rceil$) is more available than full $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness requires fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if a client's subsequent operations are executed
against a server that has observed all effects of its prior
operations, then it eventually receives a response, even in the
presence of indefinitely long partitions. As above, it may be useful
for a client to achieve stickiness \textit{and} contact many replicas
(i.e., for fault tolerance), so we generalize sticky availability to
\textit{$K$-sticky availability}, which provides sticky availability
with $K$ distinct servers. A client may choose to become $1$-sticky
available by acting as a server itself; for example, a client might
cache its reads and writes. However, a single client cannot achieve
$K$-sticky availability with $K>1$ on its own. Moreover, ``sticking''
all clients to a designated set of servers (i.e., master availability)
is a special case of sticky availability and is less available than
the equivalent non-mastered sticky availability.

@@

So far, we have focused on single-object, single-operation
availability. This is standard in distributed systems literature
(e.g., atomic, regular, and safe register models all concern single
objects~\cite{herlihy-art}), yet the database literature largely
focuses on transactions: groups of multiple operations over multiple
objects. Accordingly, by itself, replica availability is insufficient
to describe availability guarantees for transactions. Additionally,
given the choice of \textit{commit} and \textit{abort}
responses---which signal transaction success or failure to a
client---we must be careful in how we define availability.

If a client wishes to execute a transaction that performs operations
on multiple data items, then the client must be able to contact and
receive a response from at least one replica for each data item. This
may result in lower availability than a non-transactional
requirement. Additionally, given the possiblity of aborts, we need to
ensure useful forward progress. A system can trivially guarantee
clients a response by always aborting their
transactions~\cite{transaction-liveness}. However, this is an
unsatisfactory system because nothing good (transaction commit) ever
happens; instead, we require a \textit{liveness} property. We cannot
guarantee that every transaction will commit---transactions may choose
to abort themselves---but we need to make sure that the system will
not indefinitely abort transactions on its own volition. We call a
transaction abort due to a transaction's own choosing (e.g., as part
of the transaction itself or due to a would-be integrity constraint
violation) an \textit{internal abort} and an abort due to system
implementation an \textit{external abort}.

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally aborts~\cite{hat-hotos}. For example,
system will violate transactional $2$-availability if client that can
contact two servers for each of data items in its transaction does not
eventually commit in the absence of internal aborts.

>>>>

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response because the (set of) server(s) will not need to communicate
with any others. If servers are partitioned from one another, they do
not need to stall in order to provide a ``safe'' query response. This
lack of fast-path coordination also means that a highly available
system also provides low latency~\cite{abadi-pacelc}; in a wide-area
setting, this means that clients need not need to wait for
cross-datacenter communication. However, to properly describe whether
a \textit{transactional} system is highly available, we need both a
way to describe what servers a client must contact as well as what
kinds of responses a server can provide, especially given the
possibility of aborts.

@@

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage systems like
  Dynamo, we consider partially replicated systems.} a server that can
handle an operation for a given data item is called a \textit{replica}
for that item. This is a useful definition: if a client can contact at
least one replica for an item on which it wishes to perform an
operation, then the replica will provide a response to the
client. However, there are several other cases we should consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of operations will persist in the event of
server failures, clients may wish to wait until operations reach more
than one server before they consider the operations
completed. However, according to our definition of high availability,
server-fault tolerance is unavailable. To allow flexibility in the
number of participating servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing high availability ($1$-availability), a system with
$2$-availability will require more messaging and possibly higher
latency and may experience more unsuccessful operations in the
presence of network and server failures.  As is commmon in standard
consensus algorithms like Paxos, clients may want to contact a
majority of $N$ servers in a system. This \textit{majority
  availability} (a special case of $K$-availability with $K=\lceil
\frac{N}{2} \rceil$) is more available than full $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness requires fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a server that has observed all of its prior operations, it eventually
receives a response, even in the presence of indefinitely long
partitions. As above, it may be useful for a client to achieve
stickiness \textit{and} contact many replicas (i.e., for fault
tolerance), so we generalize sticky availability to \textit{$K$-sticky
  availability}, which provides sticky availability with $K$ distinct
servers. A client may choose to become $1$-sticky available by acting
as a server itself; for example, a client might cache its reads and
writes. However, a single client cannot achieve $K$-sticky
availability with $K>1$ without contacting other replicas. Moreover, ``sticking'' all clients
to a designated set of servers (i.e., master availability) is a
special case of sticky availability and is less available than the
equivalent non-mastered sticky availability.

@@

So far, we have focused on single-object, single-operation
availability. This is standard in distributed systems literature
(e.g., linearizable, regular, and safe register models all concern
single objects~\cite{herlihy-art}), yet the database literature
largely focuses on transactions: groups of multiple operations over
multiple objects. Accordingly, by itself, replica availability is
insufficient to describe availability guarantees for
transactions. Additionally, given the choice of \textit{commit} and
\textit{abort} responses---which signal transaction success or failure
to a client---we must be careful in how we define availability.

If a client wishes to execute a transaction that performs operations
on multiple data items, then the client must be able to contact and
receive a response from at least one replica for each data item. This
may result in lower availability than a non-transactional
requirement. Additionally, given the possibility of aborts, we need to
ensure that the system makes useful forward progress: a system can
trivially guarantee clients a response by always aborting all
transactions~\cite{transaction-liveness}. However, this is an
unsatisfactory system because nothing good (transaction commit) ever
happens; instead, we should require a \textit{liveness} property. We
cannot guarantee that every transaction will commit---transactions may
choose to abort themselves---but we need to make sure that the system
will not indefinitely abort transactions on its own volition. We call
a transaction abort due to a transaction's own choosing (e.g., as an
operation of the transaction itself or due to a would-be declared
integrity constraint violation) an \textit{internal abort} and an
abort due to system implementation an \textit{external abort}.

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally
aborts~\cite{hat-hotos}. For example, system will violate
transactional $2$-availability if a client that can contact two
servers for each of the data items in its transaction does not
eventually commit in the absence of internal aborts.
